---
title: 'Visualization with ggplot2: Part III'
author: "Jamison R. Crawford, MPA"
date: "May 10, 2019"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '4'
---
<style>
body {
text-align: justify}
</style>

<br>

# Review

**Base R Visualization.** For an introduction to data visualization in R, see [Visualization in Base R](http://rpubs.com/JamisonCrawford/graphics).

**Introduction to ggplot2: Part I.** See [Visualization with ggplot2: Part I](http://rpubs.com/jamisoncrawford/ggplot2), which introduces:

* The theoretical framework: "The Grammar of Graphics" (Wilkinson, 1999)
* Elements of human language vs. elements of graphical language, or layers
* "Data Ink" (**Aesthetics**) vs. "Non-Data Ink" (**Attributes**)
* Essential elements of a `ggplot2` visualization

**Introduction to ggplot2: Part II.** See [Visualization with ggplot2: Part II](http://rpubs.com/jamisoncrawford/ggplot22), which introduces:

* Introduction to most common types of plots in the **Geometries layer**
* Converting data classes to meet your visualization needs
* Common pitfalls for common **geometries**
* Implementing premade styles (**Themes**)

<br>

# Overview

This is the final introductory session on R package `ggplot2`. We'll introduce:

* Adjust scales and coordinates (the **Coordinates layer**)
* Modifying scales, labels, and other "Non-Data Ink" for clarity
* Create small multiples, i.e. trellis graphics (the **Facets layer**)
* How to overlay **geometries** for greater insights

<br>

## Practice Data

In this session, we'll use local public construction project records from:

* Construction of 2015 Lakeview Amphitheater (Source: Onondaga County)
* Renovation of 2018 Hancock International Airport (Source: Syracuse Regional Airport Authority)

You can find these datasets and documentation in this [GitHub repository](https://github.com/jamisoncrawford/wealth). These are the same data gathered by Legal Services of Central New York (LSCNY) and scraped, analyzed, and visualized for Legal Services' and Urban Jobs Task Force's collaborative publication: ["Building Equity in the Construction Trades: A Racial Equity Impact Statement"](https://issuu.com/lscny/docs/reisfinal_3-8-19_compressed).

<br>

## Required Data & Packages

Run the following code in R to install and load packages required for this session.

* Package `readr` is used for importing and exporting data
* Package `dplyr` is built around a grammar of data manipulation, inspired by SQL
* Package `scales` allows one to easily format variables, e.g. percents and dollar amounts
* Package `ggthemes`, introduced last session, expands the `ggplot2` library of preset themes

<br>

```{r warning=FALSE, message=FALSE}

if(!require("readr")){install.packages("readr")}
if(!require("dplyr")){install.packages("dplyr")}
if(!require("scales")){install.packages("scales")}
if(!require("ggplot2")){install.packages("ggplot2")}
if(!require("ggthemes")){install.packages("ggthemes")}

library(readr)
library(dplyr)
library(scales)
library(ggplot2)
library(ggthemes)

```

<br>

Run the following code to read in (i.e. import) our practice data:

* `hc` contains all scraped records for Hancock renovations
* `lv` contains all scraped records for Lakeview renovations

<br>

```{r message=FALSE, warning=FALSE}

url <- paste0("https://raw.githubusercontent.com/jamisoncrawford/",
              "wealth/master/Tidy%20Data/hancock_lakeview_tidy.csv")   # URL containing data

hc <- read_csv(file = url, 
               col_types = "ccDcccddddccliii") %>%     # Abbreviations specifying data classes
  filter(project == "Hancock")                         # Filters data for "Hancock" only

lv <- read_csv(file = url, 
               col_types = "ccDcccddddccliii") %>%     # Abbreviations specifying data classes
  filter(project == "Lakeview")                        # Filters data for "Lakeview" only

rm(url)                                                # Removes object `url`

```

<br>

# Coordinates Layer

Let's observe Hancock's worker `race` versus individual `gross` incomes per working `period`.

We're using the three essential layers for any plot, plus one additional layer for clarity:

* `ggplot()` calls the data required for the **data layer**
* `aes()` specifies which variables for the **aesthetics layer**
* `geom_jitter()` specifies a jittered scatter plot for the **geometries layer**
* `theme_light()` isn't essential, but it modifies a preset theme

<br>

```{r warning=FALSE, message=FALSE}

ggplot(hc,                            # Call data layer
       aes(x = race, 
           y = gross)) +              # Map variables in aesthetics layer
  geom_jitter(width = 0.2, 
              alpha = 0.3,
              color = "tomato") +     # Specify geometry layer, with some attributes
  theme_light()                       # Unessential: A preset theme for clarity

```

<br>

## On Coordinates

In general, the **Coordinates layer** follows a few rules:

* All functions begin with `coord_`, type this in-console to view functions via autocomplete
* Function `coord_cartesian()` is the most common and modifies a plot's Cartesian plane

<br>

## Flipping Coordinates

No, we're not politely cursing our coordinates.

Instead, we swap the x- and y-axes, which is a best practice when using categorical variables. Why?

* With many levels, categorical variables can crowd one another.
* Laypersons tend to read charts from left to right, starting in the upper-left and working down.
* In other words, general audiences will read a chart like text, in a "Z" pattern.

We can use function `ccord_flip()` to achieve this. Note that `x =` and `y=` arguments are reversed.

<br>

```{r warning=FALSE, message=FALSE}

ggplot(hc,
       aes(x = race, 
           y = gross)) +
  geom_jitter(width = 0.2, 
              alpha = 0.3,
              color = "tomato") +
  coord_flip() +                      # Swap x- and y-axes with `coord_flip()`
  theme_light()

```

<br>

## Cartesian Planes

We can "zoom in" on a part of our plot with function `coord_cartesian()`:

* Here, we'll view two continuous variables, `gross` and `net` income
* Surprisingly, these can differ since `net` includes income from other projects
* We add a Loess regression curve to show association using `geom_smooth()`
    - LOESS: Locally estimated scatterplot smoothing ([Wikipedia](https://en.wikipedia.org/wiki/Local_regression))
* First, view the plot without `coord_cartesian()`

<br>

```{r warning=FALSE, message=FALSE}

ggplot(hc,
       aes(x = net,          # Note the substitution of `race` for `net`
           y = gross)) +
  geom_point() +             # We don't need jitter and use `geom_pont()`
  geom_smooth() +            # By default, this plots a Loess curve and standard error
  theme_light()

```

<br>

Now, we'll use `coord_cartesian()` to specify where we'd like to "zoom in".

* Here, we specify the minimum and maximum zoom limits with concatenate, or `c()`
* We set the x-axis range with `xlim =` and y-axis range with `ylim =`
* Note that "zooming in" does not eliminate the other data points!

<br>

```{r warning=FALSE, message=FALSE}

ggplot(hc,
       aes(x = net,
           y = gross)) +
  geom_point() +
  geom_smooth() +
  coord_cartesian(xlim = c(0, 1000),       # We may specify limits for one or both axes
                  ylim = c(0, 2000)) +     # The range of limits in `c()` can differ
  theme_light()

```

<br>

## Filtering Planes

While `coord_cartesian()` allows one to "zoom in", limit functions will "filter" unseen data.

* Function `xlim()` with argument `x =` controls the x-axis limits
* Function `ylim()` with argument `y =` controls the y-axis limits
* Function `lims()` accepts both `x =` and `y =` arguments
* Remember to provided a minimum and maximum value in `c()`

<br>

```{r warning=FALSE, message=FALSE}

ggplot(hc,
       aes(x = net,
           y = gross)) +
  geom_point() +
  geom_smooth() +
  lims(x = c(0, 1000),
       y = c(0, 2000)) +     # Alternatively: xlim(c(0, 1000)) + ylim(c(0, 2000))
  theme_light()

```

<br>

**Questions.** Why has the Loess curve changed? Why has the standard error changed?

<br>

## Scales

Modifying the **scales** of the x- and y-axes requires you to follow a few rules, as well:

* All functions modifying scales being with `scale_`; explore with autocomplete
* You must specify the axis to modify, e.g. `scale_x_` or `scale_y_`
* You must also specify the type of variable in the function call
    - For example, if categorical e.g., use `scale_x_discrete()`
    - Conversely, if continuous e.g., use `scale_y_continuous()`
* The following modifies x- and y-axes labels, or `labels =`
    
<br>

```{r warning=FALSE, message=FALSE}

ggplot(hc,
       aes(x = net,
           y = gross)) +
  geom_point() +
  geom_smooth() +
  scale_x_continuous(name = "Unique Net Payments") +       # Specify continuous x-axis
  scale_y_continuous(name = "Unique Gross Payments") +     # Specify continuous y-axis
  theme_light()

```

<br>

**Labels.** Above, we've simply modifed the axes labels with argument `name =`.

We can use R package `scales` to easily format the data labels, which may:

* Add comma separators for values of 1,000 or more (`label = "comma"`)
* Convert to dollars (`labels = dollar`) or percents (`labels = percent`)

<br>

```{r warning=FALSE, message=FALSE}

library(scales)

ggplot(hc,
       aes(x = net,
           y = gross)) +
  geom_point() +
  geom_smooth() +
  scale_x_continuous(name = "Unique Net Payments",
                     labels = dollar) +             # `scales` makes formatting easy!
  scale_y_continuous(name = "Unique Gross Payments",
                     labels = dollar) +
  theme_light()

```

<br>

**Breaks.** We can also format **breaks** along each axis, i.e. labels every $500, or every $1K.

* These breaks are numeric and so do not require quotiation marks
* Specify breaks using the argument `breaks =`
* Individual breaks are concatenated in `c()`

<br>

```{r warning=FALSE, message=FALSE}

library(scales)

ggplot(hc,
       aes(x = net,
           y = gross)) +
  geom_point() +
  geom_smooth() +
  scale_x_continuous(name = "Unique Net Payments",
                     labels = dollar, 
                     breaks = c(0, 500, 1000, 1500, 2000, 2500)) +     # Custom breaks
  scale_y_continuous(name = "Unique Gross Payments",
                     labels = dollar,
                     breaks = c(0, 500, 1000, 1500, 2000, 2500, 3000, 3500)) +
  theme_light()

```

<br>

**Repeating Breaks.** We don't need to type out each "break" value by using package `scales`:

* Function `pretty_breaks()` from `scales` allows easy, incremental break formatting
* Argument `n =` specifies the numebr of equal breaks

<br>

```{r warning=FALSE, message=FALSE}

library(scales)

ggplot(hc,
       aes(x = net,
           y = gross)) +
  geom_point() +
  geom_smooth() +
  scale_x_continuous(name = "Unique Net Payments",
                     labels = dollar, 
                     breaks = pretty_breaks(n = 6)) +     # Pretty breaks
  scale_y_continuous(name = "Unique Gross Payments",
                     labels = dollar,
                     breaks = pretty_breaks(n = 8)) +
  theme_light()

```

<br>

**Custom Labels.** We don't have to use package `scales`, though it helps. Instead:

* We can create x- and y-axis labels that use non-data ink more parsimoniously
* This allows audiences to spend more time focusing on the data
* However, we need to provide sufficient context with some "data ink"
* The number of unique labels **must match** number of breaks!
* Here, by incuding "$", we need not include "USD" in labels

<br>

```{r warning=FALSE, message=FALSE}

ggplot(hc,
       aes(x = net,
           y = gross)) +
  geom_point() +
  geom_smooth() +
  scale_x_continuous(name = "Unique Net Payments",
                     labels = c("$ 0", "0.5", "1.0", "1.5", "2.0", "2.5 K"),
                     breaks = c(0, 500, 1000, 1500, 2000, 2500)) +
  scale_y_continuous(name = "Unique Gross Payments",
                     labels = c("$ 0", "0.5", "1.0", "1.5", "2.0", "2.5", "3 K"),
                     breaks = c(0, 500, 1000, 1500, 2000, 2500, 3000)) +
  theme_light()

```

<br>

# Faceting

[Small multiples](https://en.wikipedia.org/wiki/Small_multiple) allow us to compare different groups of data using the same scales.

* Also known as faceting or trellis graphics, this is the **Facets layer**
* All faceting functions begin with `facet_`; peruse faceting with autocomplete
* Typically, functions `facet_grid()` and `facet_wrap()` are most common
    - `facet_grid()` aligns small multiples horizontally along one shared y-axis
    - `facet_wrap()` structures small multiples tabularly, in rows and columns
* Facet functions accept a formula with bare variable names, e.g. `class ~ race`
    - If you prefer to facet by one variable, leave one side empty, or use `.`
    - The lefthand-side specifies faceting rows, and RHS specifies columns
    
Let's try `facet_grid()` first with formula `~ race`, faceting on `race` alone:
    
<br>

```{r warning=FALSE, message=FALSE}

ggplot(hc,
       aes(x = net,
           y = gross)) +
  geom_point() +
  geom_smooth(method = "lm") +                           # Linear model, not Loess
  scale_x_continuous(name = "Unique Net Payments",
                     labels = c("$0", "1", "2 K", "")) +
  scale_y_continuous(name = "Unique Gross Payments",
                     labels = dollar) +
  facet_grid( ~ race) +                                  # Use a single y-axis scale 
  theme_light() 

```

<br>

Now let's give `facet_wrap()` a try with the same formula:

* This tends to work better when faceting by 5 or more groups

<br>

```{r warning=FALSE, message=FALSE}

ggplot(hc,
       aes(x = net,
           y = gross)) +
  geom_point() +
  geom_smooth(method = "lm") +                           # Linear model, not Loess
  scale_x_continuous(name = "Unique Net Payments",
                     labels = c("$0", "1", "2 K", "")) +
  scale_y_continuous(name = "Unique Gross Payments",
                     labels = dollar) +
  facet_wrap( ~ race) +                                  # Use a tabular format 
  theme_light() 

```

<br>

There are several arguments for faceting functions, though usually you only need a formula.

<br>

# Labels

Labeling is also fairly straightforward and usually contain `lab` in the function name:

* Function `labs()` is a catch-all for multiple labels
* You can modify labels elsehow, e.g. in `scale_x_continuous()`
* Alternatively, individual functions exist for modifying each label:
    - `ggtitle()` modifies the plot title
    - `xlab()` modifies the x-axis title
    - `ylab()` modifies the y-axis title
    - Etc.
    
Most conveniently, you can modify all labels in the `labs()` function using arguments:

* `title =` for plot title
* `subtitle = ` for plot subtitle
* `x =` for x-axis title
* `y =` for y-axis title
* `caption =` for listing sources
* `fill = `, `color =`, or other aesthetics for legend titles

<br>

```{r warning=FALSE, message=FALSE}

ggplot(hc,
       aes(x = net,
           y = gross)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_x_continuous(labels = dollar) +
  scale_y_continuous(labels = dollar) +
  facet_wrap( ~ race) +
  labs(title = "Gross vs. Net payments by ethnicity",                 # Labels in fell swoop
       subtitle = "2018 Hancock Airport Renovations",
       x = "Unique Net Payments",
       y = "Unique Gross Payments",
       caption = "Source: Syracuse Regional Airport Authority") +
  theme_light() 

```

<br>

# Plot Overlays

**Overlay** plots by adding new `geom_*()` and `aes()` functions and arguments appropriately.

* Here, we'll go back to our `geom_jitter()` plot with `gross` income and `race`
* If you're using new data for your overlay, specify it in the `geom_*()` call
* If you're using new variables, or differently named variables, specify in `aes()`

<br>

```{r warning=FALSE, message=FALSE}

ggplot(hc,
       aes(x = race,
           y = gross)) +
  geom_boxplot(width = 0.55,
               color = "grey55") +
  geom_jitter(width = 0.2,
              alpha = 0.3,
              color = "tomato") +
  coord_flip() +
  scale_y_continuous(labels = dollar) +
  labs(title = "Gross payments by ethnicity",
       subtitle = "2018 Hancock Airport Renovations",
       x = "Ethnicity",
       y = "Unique Gross Payments",
       caption = "Source: Syracuse Regional Airport Authority") +
  theme_light() 

```

<br>

# Applied Practice

Instead of dataset `hc`, for Hancock data, replicate the above plots with Lakeview data, `lv`. Recommendations are provided below, but feel free to experiment. Make sure to:

* Call dataset `lv` in your data layer, `ggplot()`
* Use the same variable, which you can view with `names(lv)`
* Label your titles, subtitles, axes, legends, and captions appropriately
* Format your scales approrpiately, if dollars, percents, etc.
* Use non-data ink parsimoniously and for clarity

<br>

Suggested plots:

1. A `geom_jitter()` plot with `race` by `gross`
2. A faceted `geom_point()` plot with `geom_smooth()` using `gross` by `net`
3. A `geom_bar()` plot with argument `stat = "identity"` in the geometry layer
    - Use `fill = race` in the `aes()` function call
    
<br>

# Acknowledgments

Thanks for sticking it out with me throughout this introductory series to the R language.

With grit and determination, which you've already shown, these fundamentals will be a robust foundation, without pretense to architecture, on which to build your future data science skills. We've only scratched the surface - tooth and nail, of course - but such is the nature of any aspiring hacker. Keep it up.

In the words of Benjamin Franklin:

> "Industry need not wish".