---
title: 'Intro to R: Text Data'
author: "Jamison R. Crawford, MPA"
date: "November 8, 2018"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    number_sections: true
---

```{r echo=FALSE, message=FALSE, warning=FALSE}

setwd("~/CNYCF/R Learning Community/2018-11-09")

```

**Latest Versions & Updates:** This markdown document was built using the following versions of *R* and *RStudio*:
 
* R v. 3.5.1
* RStudio v. 1.1.456
* Document v. 1.0
* Last Updated: 2018-11-08

<br>

# Review

The following provides an overview of techniques we've learned, including links to the original session.

<br>

## Objects & Assignment

**Objects** in R contain single values, multiple values (*vectors*), and tabular data (*data frames*). 
 
**The Assignment Operator**, `<-`, names and stores one or more values, functions, or data structures.

```{r}
my_value <- 5                                            # Store a single value

my_vector <- c(5, 10, 15)                                # Vectors: Concatenated values

my_dataframe <- data.frame(x = c(1, 2, 3),
                           y = c("a", "b", "c"),
                           z = c(TRUE, TRUE, FALSE))     # Data Frames: Tabular structures
```

<br>

**Print objects** by simply entering the object name or explicitly using the function `print()`.

```{r}
my_value             # Autoprints using only the object name
```

```{r}
print(my_vector)     # Explicitly prints with function print() 
```

<br>

**Built-In Objects** already exist in R, such as `letters`, all lowercase letters, or `mtcars`, a dataset on cars from 1972.

```{r}
letters
head(mtcars)
```

**Original Session****:** [Intro to R: Operators](http://rpubs.com/JamisonCrawford/operators)

<br>

## Operators

**Arithmetic Operators** in R are used for addition, subtraction, multiplication, division, operator preference, and exponentiation.

**Class Numeric** data are required.

```{r}
(5^2 * 4) / 2
```

<br>

**Relational Operators** in R are used in relational statements that compare one or a series of values, e.g. `<`, `>`, `==`, `!=`.

**Class Logical** result from relational statements, i.e. `TRUE` or `FALSE`. 

```{r}
10 < c(8, 9, 11, 12)
```

<br>

**Logical Operators** bind multiple relational statements.

**OR**, i.e. `|`, requires at least one statement to be `TRUE`.

```{r}
5 > 1 | 10 < 5
```

**AND**, i.e. `&`, requires all statements to be `TRUE.

```{r}
5 > 1 & 10 < 5
```

**Original Session****:** [Intro to R: Operators](http://rpubs.com/JamisonCrawford/operators)

<br>

## Subsetting & Indexing

**The Dollar Sign Operator**, i.e. `$`, subsets or extracts a specific variable from a dataset.

```{r}
mtcars$mpg          # Combine the dataset name and variable name to subset the variable
```

<br>

**Indexing** a subset variable is done with brackets, `[` & `]`, and the number or numbers of the element(s) by position.

```{r}
mtcars$mpg[5]      # Combine the dataset, variable, and position to extract a specific value
```

<br>

**Index by Row & Column Position** using the row number and column number in brackets, separated by a comma, `,`.

```{r}
mtcars[25, 1]
```

<br>

**Index by Name** using the row name and column name in the same manner.

```{r}
mtcars["Pontiac Firebird", "mpg"]
```

<br>

**Index Multiple Positions** by concatenating more than one position number using function `c()`.

```{r}
mtcars["Honda Civic", c(1, 2, 4, 6)]
```

<br>

**Subset All Rows or All Columns** by leaving the position empty within the brackets.

```{r}
mtcars[1:5, ]        # Subset rows 1-5 and all columns
mtcars[ c(1, 2)]     # Subset columns 1-2 and all rows
```

<br>

**Filter with Relational Operators** by placing a relational statement in the row position, in brackets.

```{r}
mtcars[mtcars$mpg < 15, ]     # Subset only cars with less than 15 mpg
```

<br>

**Assign Subset Data to New Objects** using the assignment operator, `<-`, an object name, and the subset data.

```{r}
gas_guzzlers <- mtcars[mtcars$mpg < 15, ]
```

<br>

**Save Objects to Index Data** using the assignment operator, `<-` and one or more relational statements.

```{r}
index <- mtcars$cyl == 8 & mtcars$hp > 240     # Store logical values: TRUE or FALSE

dream_cars <- mtcars[index, ]                  # Use the indexing object in the row position

print(dream_cars)                              # Print results
```

**Original Session****:** [Intro to R: Subsets & Indices](http://rpubs.com/JamisonCrawford/subsets)

<br>

# On Classes

**Classes** of both variables and single values dictate how R will recognize and work with them.

<br>

## Determining Class

**Identify Class** by using the `class()` function and inputting either one or more values or an object.

```{r}
class(10L)                # Call class() on a single value; here, "L" indicates an integer
class(c(TRUE, FALSE))     # Call class() on multiple values, e.g. "logical" values
class(mtcars)             # Call class() on an object with stored data to determine structure
class(mtcars$mpg)         # Call class() on a subset variable for the class of its values
```

<br>

**Numeric** data include any quantitative data, including:

* Class `numeric` in an all encompassing term for quantitative data
* Class `integer`, or values comprised of whole numbers
* Class `double`, or values with floating decimals

<br>

**Logical** data contain logical values, e.g. `TRUE` or `FALSE`.

Under the hood, `logical` data are represented by 1 and 0.

```{r}
TRUE == 1
FALSE == 0
```

<br>

**Character** data contain uncategorized text, e.g. "Onondaga County".

```{r}
my_county <- "Onondaga County"
class(my_county)
```

<br>

**Factor** data represent categorical data where each category is a "level", e.g. gender, race, or census tract.

```{r}
cylinders <- factor(mtcars$cyl)        # Create factors using the factor() function
class(cylinders)
levels(cylinders)                      # Function levels() prints each category in a factor
```

<br>

## Coercing Classes

**Coercion** is the act of converting values and objects to new classes, usually with an `as.()` function.

```{r}
class(mtcars$mpg)                          # Print the class of variable "mpg"
mtcars$mpg <- as.character(mtcars$mpg)     # Coerce the class from "numeric" to "character"
class(mtcars$mpg)                          # Re-print the class to confirm changes
```

<br>

**The Purpose of Coercion** is so R will treat your values in the manner you intend.

**Function Overloading** is the quality in R which allows functions to behave differently depending on object class.

```{r}
class(mtcars$cyl)       # Determine class for variable "cyl", or number of cylinders
summary(mtcars$cyl)     # Print descriptive statitistics for numeric data with summary()
```

```{r}
mtcars$cyl <- as.character(mtcars$cyl)     # Coerce to class "character" with as.character()
summary(mtcars$cyl)                        # Function summary() now prints the number of elements
```

```{r}
mtcars$cyl <- as.factor(mtcars$cyl)     # Coerce to class "factor" with as.factor()
summary(mtcars$cyl)                     # Prints each "level" (category) and frequency of each
```

<br>

**Identify All Classes in a Dataset** by using the function `str()`, or "structure", which prints the:

* Dimensions of the dataset, i.e. total rows (observations) and columns (variables)
* Class of the data structure
* Class of each variable
* First few values of each variable
* Quantitaty and name of each factor level

```{r}
str(iris)     # Print the structure of the "iris" dataset, or 150 measures of iris species
```

<br>

## Importance of Coercion

**Coercion in Data Visualization** is also very important. Observe the following. What do you notice about the x-axis?

```{r}
data(mtcars)
plot(x = mtcars$cyl, 
     y = mtcars$mpg,
     col = "tomato",
     xlab = "Number of Cylinders",
     ylab = "Miles per Gallon",
     main = "Cylinders vs. MPG")
```

**Note****:** R identifies two continuous variables and makes a scatterplot, assuming 5- and 7-cylinder engines are missing.

<br>

**Prevent Categorical Variables from Appearing Continuous** by coercing "numeric" variables to class "factor". 

```{r}
data(mtcars)
plot(x = as.factor(mtcars$cyl),        # The only change is nesting the variable in as.factor()
     y = mtcars$mpg, 
     col = "tomato",
     xlab = "Number of Cylinders",
     ylab = "Miles per Gallon",
     main = "Cylinders vs. MPG")
```

**Function Overloading** occurs as function `plot()` now acknowledges the "factor", creating a box plot.

<br>

**Coercion in Regression** is even more important. Using `lm()`, we'll try to create a linear model with the original `mtcars`.

```{r}
data(mtcars)
my_lm <- lm(mpg ~ cyl, 
            data = mtcars)
print(my_lm)
```

**Incorrect Interpretation:** Per the coefficients, every unit of `cyl` added reduces `mpg` by 2.87. This is absurd.

<br>

```{r}
data(mtcars)
my_lm <- lm(mpg ~ as.factor(cyl), 
            data = mtcars)
print(my_lm)
```

**Correct Interpretation****:** 6-cylinder engines reduce `mpg` by 6.92, while 8-cylinder engines reduce `mpg` by 11.56.

<br>

## Practice

**Coercing Numeric Data** is equally important, as demonstrated in the following scenario.

**Scenario****:** You're colleague has written a PDF scraper to extract key Form 990 data, seen in dataset `form_990`:

```{r}
form_990 <- data.frame("FY_2017" = c("764882", "240739", "49212"), 
                       "FY_2018" = c("841912", "263997", "41315"), 
                       stringsAsFactors = FALSE, 
                       row.names = c("Programming Expenses", 
                                     "Administrative Expenses", 
                                     "Fund Development Expenses"))
print(form_990)
```

**Practice****:** Find the sum total of all expenses in fiscal years 2017 and 2018.

* Determine the classe of each variable using `class()`.
* Subset each variable in `form_990` using the `$` operator.
* Use the appropriate coercion function to render the data usable.
* Use function `sum()` to find the total of each fiscal year.
* Use function `sum()` again on the totals.

<br>

**Conclusions****:** Identifying variable classes is a crucial first step in exploratory data analysis. As demonstrated above, failing to identify and coerce classes can be fatal to the accuracy of your analyses and visualizations. We've only looked at coercion with "numeric" and "factor" classes, but for nearly every data class (and there are many more), there is a way to coerce it to a more appropriate and actionable class.

<br>

## Further Resources: Factors

**Learn More** about `factor()` and `as.factor()` by calling `help(factor)` and `help(as.factor)` within R. In addition, I highly recommend exploring the fourth module in DataCamp's free [Introduction to R](https://www.datacamp.com/courses/free-introduction-to-r).

<br>

# Text Data

The following provides an overview of base R functions for data of class "character". Run the following in R.

```{r}
url <- "https://tinyurl.com/y9xuc5pa"
construct <- read.csv(file = url, stringsAsFactors = FALSE); rm(url)
```

These are the records of *Quality Structures, Inc.*, the largest of multiple contractors working on Syracuse International Airport's 2018 renovations and retrieved via Freedom of Information Act (FOIA).

Read the documentation here: [REIS GitHub Repository](https://github.com/jamisoncrawford/REIS). 

<br>

## An Introduction

**Overview****:** Data of class "character" is often easily distinguishable due to quotations, e.g. `"this"`. 

Any values you write or store are automatically converted to class "character" when using quotations. Observe:

```{r}
my_word <- "perspicacity"     # Quotes guarantee that value will be stored as class "character"
class(my_word)
print(my_word)
```

<br>

**String Manipulation** is the act of manipulating text data, most often referred to as **strings**.

We can think of "strings" as a *sequence of characters*, which may be alphabetical or numeric.

<br>

## Pasting Strings

**Pasting** is the act of combining multiple strings to form a longer or more complex string, performed with `paste()`.

```{r}
x <- "I'm"
y <- "learning"
z <- "R!"
paste(x, y, z, sep = " ")     # Argument "sep =" specifies the character between pasted strings
```

<br>

**Notice that** this we've pasted together objects, but you can just as easily input the strings by hand:

```{r}
paste("I'm", "learning", "R!", sep = " ")
```

<br>

**The versatility** of `paste()` is often underappreciated at first glance. We could goof off by tampering with `sep =`:

```{r}
paste("Millennial:", x, y, z, sep = ", like, ")
```

<br>

**We could** do something more useful, like combine names in a character roster. First, let's create one:

```{r}
first <- c("Luis", "Cody", "Shannon", "Jamison")
last <- c("Escoboza", "Peck", "Connor", "Crawford")
roster <- data.frame(first, last)
print(roster)
```

<br>

**Now** we can use paste to make a "Surname, First Name" format, like so:

```{r}
paste(roster$last, roster$first, sep = ", ")

```

<br>

**Now we can add it** as a new variable using `$`.

```{r}
roster$both <- paste(roster$last, roster$first, sep = ", ")
print(roster)
```

<br>

**We could also** create a sequence of URLs for a web crawler, e.g. adult literacy programs around Dallas, TX:

```{r}
url <- "https://www.nationalliteracydirectory.org/programs?q=75201&radius=25&student="
iteration <- as.character(c(1:6))

paste(url, iteration, sep = "")
```

<br>

**Conclusions:** The *National Literacy Directory*, which provides the search results for the above pages, is owned by the *Dollar General Literacy Foundation* - and they absolutely do not want those data shared, despite the tremendous potential it could have in the hands of researchers for ameliorating today's adult literacy crisis. Fortunately, you can only search within a 25-mile radius, which limits the amount of search options.

Of course, if you just change `radius=25` in `url` to, say, `radius=6000`, or 1/4 of the circumference of the earth, you'd have every adult education program in the United States, including Hawaii and Alaska. That's 537 individual pages of search results through which one could sequence, covering 10,730 programs. *But you should definitely not do that*.

In sum, `paste()` is extremely useful. Never forget it.



